Effects of epidemiological structure on the transient evolution of HIV virulence

### 2016-10-18 ms/plos_latex.tex
tweaking ms
add latexdiff comparison

\end{flushleft}
% Please keep the abstract below 300 words
% current is 298!
\section*{Abstract}
The evolutionary dynamics of parasite virulence over the
  course of an emerging epidemic have important implications both for
  our basic understanding of epidemiological dynamics and,
  potentially, for the outcomes of public health interventions.
  Changes in the fitness landscape
  will generally select for higher virulence during the early
  phase of an epidemic, but quantitative outcomes
  may depend sensitively on biological details and the structure of
  mathematical models used to capture them.  Fraser and
  co-workers have proposed a series of models for eco-evolutionary
  dynamics of HIV that are relatively detailed in their portrayal of
  the tradeoffs between transmission and virulence (mediated by
  set-point viral load, SPVL) and their heritability between
  hosts. However, these models use implicit
  representations of the transmission process that drastically simplify the
  partnership dynamics that previous research has found to be critical
  in driving epidemics of sexually transmitted diseases.  We
  explore models that combine HIV virulence tradeoffs with a range of
  epidemiological structures, modeling partnership formation and
  dissolution and allowing for individuals to transmit disease outside
  of partnerships. We assess summary statistics such as the minimum
  time of progression to AIDS and the time at which this 
  extreme value (peak virulence) occurs across all
  models for a range of 
  partnership dynamic parameters 
  applicable to the HIV epidemic in sub-Saharan Africa. In
  order to account for the different interpretations of parameters
  across model structures, we scale all parameter sets to constrain
  the simulated epidemic growth rate to be identical, matching a
  realistic baseline value. 
  Although virulence trajectories are broadly similar
  across model structures --- an initial increase in virulence/decrease
  in progression time, followed by a minimum and an increase to an equilibrium 
  progression time slightly higher than the minimum --- the timing and magnitude of the minimum vary
  considerably.
  Models of intermediate
  complexity predicted much lower peak virulence (15 years to
  progress to AIDS) compared to both more realistic models
  and simple random-mixing models with no partnership structure
  at all (both approx. 7.25 years to progress to AIDS).
  In this range of models, the simplest random-mixing structure is
  actually the best approximation to the most realistic model; this
  surprising outcome occurs because the dominance of extra-pair
  contact in the realistic model tends to mask the effects of
  partnership structure.

@@

Lastly, we add a model of heterogeneity in sexual activity to the pairform+epc model (``hetero''). Individuals are divided into different risk groups based on the sexual activity level; we scale all aspects of sexual activity, assuming that sexual activity level in both within- and extra-couple contacts is directly proportional to number of non-cohabiting (extra-couple and uncoupled) partners per year \cite{omori2015dynamics} (see Appendix for model details). We assume random activity-weighted mixing between risk groups \todo{cite old Anderson-May HIV paper?}. While this model lacks some 
important elements, such as age-structured mixing patterns, needed for realistic models of HIV transmission in sub-Saharan Africa, we see it as a first step toward assessing the effects of epidemiological complexity. As even these models are pushing the limits of compartmental-based models, adding further complexity will probably require a shift to an agent-based model framework, as well as considerable effort in model calibration \cite{herbeck_hiv_2014,delva_connecting_2016}.

>>>>

\end{flushleft}
% Please keep the abstract below 300 words
% current is 298!
\section*{Abstract}
The evolutionary dynamics of parasite virulence over the
course of an emerging epidemic have important implications for
our understanding of epidemiological dynamics.
While changes in the fitness landscape
generally select for higher virulence early in an
epidemic, quantitative outcomes
may depend sensitively on biological details and the structure of
mathematical models used to portray them.  Fraser \emph{et al.} 
have proposed a model for the eco-evolutionary
dynamics of HIV that captures
the tradeoffs between transmission and virulence (mediated by
set-point viral load, SPVL) and their heritability between
hosts. However, these models use implicit
representations of the transmission process that drastically simplify the
partnership dynamics that previous research has found to be critical
in driving epidemics of sexually transmitted diseases.  
Our models combine HIV virulence tradeoffs with a range of
epidemiological structures, modeling partnership formation and
dissolution and allowing for individuals to transmit disease outside
of partnerships. We assess summary statistics such as the peak virulence
(corresponding to the minimum expected time of progression to AIDS) across all
models for a range of 
partnership dynamic parameters 
applicable to the HIV epidemic in sub-Saharan Africa.
Although virulence trajectories are broadly similar
across model structures, the timing and magnitude of the 
minimum expected time to progression vary
considerably.
Models of intermediate
complexity predicted slower progression (a minimum of 15 years to
progress to AIDS) compared to both more realistic models
and simple random-mixing models with no partnership structure
at all (both with a minimum of $\approx$ 7.25 years to progress to AIDS).
In this range of models, the simplest random-mixing structure best
approximates the most realistic model; this
surprising outcome occurs because the dominance of extra-pair
contact in the realistic model tends to mask the effects of
partnership structure.

@@

Lastly, we add a model of heterogeneity in sexual activity to the pairform+epc model (``hetero''). Individuals are divided into different risk groups based on the sexual activity level; we scale all aspects of sexual activity, assuming that sexual activity level in both within- and extra-couple contacts is directly proportional to number of non-cohabiting (extra-couple and uncoupled) partners per year \cite{omori2015dynamics} (see Appendix for model details). We assume random activity-weighted mixing between risk groups \cite{may_transmission_1988}. While this model lacks some 
important elements, such as age-structured mixing patterns, needed for realistic models of HIV transmission in sub-Saharan Africa, we see it as a first step toward assessing the effects of epidemiological complexity. As even these models are pushing the limits of compartmental-based models, adding further complexity will probably require a shift to an agent-based model framework, as well as considerable effort in model calibration \cite{herbeck_hiv_2014,delva_connecting_2016}.
